\renewcommand{\bibname}{Notes \&\ References}
\begin{thebibliography}{999}

\addcontentsline{toc}{chapter}{Notes \&\ References}

\bibitem{TapFormat} 
  \textbf{TAP Format},
  \href{http://unusedino.de/ec64/technical/formats/tap.html}{A description of the TAP Format},
  \href{https://www.c64-wiki.com/wiki/TAP}{C64 Wiki}.
\bibitem{MegaSave} 
  \textbf{The Megasave Loader},
  \href{https://www.luigidifraia.com/doku/doku.php?id=commodore:tapes:loaders:mega-save}{Luigi Di Fraia's Mega SaveLoader page}.
\bibitem{LlamasoftHistory} 
  \textbf{A History of Llamasoft},
  \href{http://minotaurproject.co.uk/YakImages/AHistoryofLlamasoft.pdf}{Jeff Minter's memoir of the early years of Llamasoft}.
\bibitem{c64prg} 
  \textbf{C64 Programmer's Reference Guide},
  \href{https://archive.org/details/c64-programmer-ref}{The bible for C64 programmers}.
\bibitem{6502} 
  \textbf{Programming the 6502}, Rodnay Zaks
  \href{https://archive.org/details/Programming_the_6502_OCR}{The book used by Jeff Minter to learn assembly.} "Learning Machine Code ... I used PROGRAMMING THE 6502 by Rodney Zaks as a good 6502 book, and used the VIC and 64 Programmers Reference Guides for each machine. The VIC REVEALED is also good for VIC 20 owners."
\bibitem{ia} 
  \textbf{Iridis Alpha Source Code},
  \href{https://github.com/mwenge/iridisalpha}{Source code for Iridis Alpha, reverse-engineered by the author of this book.}
\bibitem{iridisdownload} 
  \textbf{Iridis Alpha: Tape Download},
  \href{http://www.llamasoftarchive.org/oldsite/llamasoft/cbm64/IridisAlpha.zip}{A tape copy of Iridis Alpha.}

\bibitem{deflex} 
  \textbf{The Origins of Deflex},
  \href{https://web.archive.org/web/20230811235533/https://stinkygoat.livejournal.com/211183.html}{Jeff Minter's history of Deflex.}

\bibitem{devdiary} 
  \textbf{Iridis Alpha Development Diary},
    \href{https://web.archive.org/web/20230322100622/https://codetapper.com/c64/diary-of-a-game/iridis-alpha/}{Jeff Minter's Development Diary, published in Zzap! magazine from issue 13 (May 1986) to issue 17 (September 1986).}

\bibitem{beast} 
  \textbf{The Nature of the Beast},
    \href{https://web.archive.org/web/20230322100622/https://codetapper.com/c64/diary-of-a-game/iridis-alpha/}{ Jeff Minter's regular newsletter to Llamasoft subscribers.}
\end{thebibliography}
